---
title: "Batch correction for the droplet data"
author: "Aaron Lun, based on code by Laleh Haghverdi and Michael Morgan"
date: "Revised: 28 February 2019"
output:
  BiocStyle::html_document:
    toc_float: true
    fig_caption: no
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library(BiocStyle)
library(BiocParallel)
register(SerialParam())
```

# Overview

Here we test out `fastMNN` on the droplet data.
In the interests of time, we won't bother testing the others.
First, we normalize both batches to the same scale.

```{r}
library(SingleCellExperiment)
sce.68 <- readRDS("sce.pbmc68k.rds")
sce.4 <- readRDS("sce.t4k.rds")
universe <- intersect(rownames(sce.68), rownames(sce.4))

library(batchelor)
normed <- multiBatchNorm(sce.68[universe,], sce.4[universe,])
sce.68 <- normed[[1]]
sce.4 <- normed[[2]]
```

Taking all genes with positive biological components.

```{r}
dec.68 <- readRDS("dec.pbmc68k.rds")
dec.4 <- readRDS("dec.t4k.rds")
combined <- (dec.68[universe,"bio"] + dec.4[universe,"bio"])/2
to.use <- combined > 0
summary(to.use)
```

Subsetting to these genes.

```{r}
sce.68 <- sce.68[to.use,]
sce.4 <- sce.4[to.use,]
```

```{r, echo=FALSE, results="hide"}
gc()
```

# Setting up the visualization

Organizing colors for cell types and batches.

```{r}
batch.id <- rep(1:2, c(ncol(sce.68), ncol(sce.4)))
batchcolor <- c("lavender", "lightcoral")[batch.id]

cd3e <- c(logcounts(sce.68)["ENSG00000198851",], logcounts(sce.4)["ENSG00000198851",])
cd3e <- pmin(cd3e, 2) # to preserve dynamic range.
allcolors <- viridis::viridis(20)[cut(cd3e, 20)]
```

Making a plotting function for each metric.

```{r}
plotFUN <- function(fname, Y, subset=NULL, ..., xlab="tSNE 1",ylab="tSNE 2",main="") {
    if (is.null(subset)) {
        subset <- seq_len(nrow(Y))
    }
    png(fname,width=900,height=700)
    par(mfrow=c(1,1),mar=c(6,6,4,2),cex.axis=2,cex.main=3,cex.lab=2.5)
    plot(Y[,1], Y[,2], cex=2, pch=21, bg=allcolors[subset], xlab=xlab, ylab=ylab, main=main) 
    dev.off()
}

plotFUNb <- function(fname, Y, subset=NULL, ...) {
    if (is.null(subset)) {
        subset <- seq_len(nrow(Y))
    }
    png(fname,width=900,height=700)
    par(mfrow=c(1,1),mar=c(6,6,4,2),cex.axis=2,cex.main=3,cex.lab=2.5)
    plot(Y[,1], Y[,2], cex=2, pch=21, bg=batchcolor[subset], ...)
    dev.off()
}
```

# No correction

Running it without any correction (but with a PCA, to make things manageable w.r.t. time).

```{r}
library(BiocSingular)
set.seed(1000)
out <- multiBatchPCA(logcounts(sce.68), logcounts(sce.4), d=50, 
    BSPARAM=IrlbaParam(deferred=TRUE))
t.unc <- do.call(rbind, out)
```

Generating a t-SNE plot.

```{r}
set.seed(0)
library(Rtsne)
tsne.unc <- Rtsne(t.unc, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_unc_type.png", tsne.unc$Y, main="Uncorrected", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_unc_batch.png", tsne.unc$Y, main="Uncorrected", xlab="tSNE 1",ylab="tSNE 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.unc)
gc()
```

# Using fast MNN

Performing the correction with faster MNN.

```{r}
set.seed(1000)
mnn.out2 <- fastMNN(logcounts(sce.68), logcounts(sce.4), k=20, 
    BSPARAM=IrlbaParam(deferred=TRUE))
t.mnn <- reducedDim(mnn.out2)
```

Generating a t-SNE plot.

```{r}
set.seed(0)
tsne.mnn <- Rtsne(t.mnn, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_mnn2_type.png", tsne.mnn$Y, main="Fast MNN", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_mnn2_batch.png", tsne.mnn$Y, main="Fast MNN", xlab="tSNE 1",ylab="tSNE 2")
```

# Session information

```{r}
sessionInfo()
```
