---
title: "Batch correction for the haematopoiesis data sets"
author: "Aaron Lun, based on code by Laleh Haghverdi and Michael Morgan"
date: "Revised: 27 February 2019"
output:
  BiocStyle::html_document:
    toc_float: true
    fig_caption: no
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview

This script performs batch correction with each method and creates t-SNE plots for the haematopoietic data sets.
First we load in the data:

```{r}
library(SingleCellExperiment)
dir.create("results", showWarning=FALSE)
sceA <- readRDS("haem_data_A.rds")
sceF <- readRDS("haem_data_F.rds")
```

We re-normalize to bring the two batches onto the same scale.
This requires us to only use the genes that are present in both batches. 

```{r}
library(batchelor)
common <- intersect(rownames(sceA), rownames(sceF))
nout <- multiBatchNorm(sceA[common,], sceF[common,])
sceA <- nout[[1]]
sceF <- nout[[2]]
```

We set up the file cache and the seed for later use.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache(ask=FALSE)    
set.seed(123456)
```

# Identify highly variable genes

We use an existing list of highly variable genes identified by Nestrowa _et al._
It is straightforward to do this empirically but we will save ourselves the trouble.

```{r}
hvg.fname <- bfcrpath(bfc,
    "http://blood.stemcells.cam.ac.uk/data/coordinates_gene_counts_flow_cytometry.txt.gz")
TFs <- read.table(hvg.fname, nrows=1, stringsAsFactors=FALSE)
features <- as.character(unlist(TFs))
features <- features[grep("ENSMUS", features)]
length(features)
```

We obviously only retain the genes that are present in our two objects.

```{r}
stopifnot(identical(rownames(sceA), rownames(sceF)))
keep <- rownames(sceA) %in% features
sceA <- sceA[keep,]
sceF <- sceF[keep,]
summary(keep)
```

# Setting up visualizations

First, we need to match the annotation types between the two objects.

```{r}
sceF$CellType[sceF$CellType=="other"] <- "Unsorted"
table(sceF$CellType)
sceA$CellType[sceA$CellType=="ERY"] <- "MEP"
table(sceA$CellType)
```

We now combine them together:

```{r}
sce <- cbind(sceF, sceA)
sce$Batch <- rep(c("Nestorowa", "Paul"), c(ncol(sceF), ncol(sceA)))
```

We define colours for each cell type in each batch.
Basically, the Paul data set uses a lighter shade of colours.

```{r}
color.legendF <- c(MEP="orange", GMP="chartreuse4", CMP="magenta", 
    HSPC="cyan", LTHSC="dodgerblue", MPP="blue", LMPP="light blue", Unsorted="grey")
colmatF <- col2rgb(color.legendF) 
colmatA <- colmatF + 100 # A lighter shade.
colmatA[colmatA > 255] <- 255

color.legendA <- setNames(rgb(colmatA[1,], colmatA[2,], colmatA[3,], maxColorValue=255), 
    names(color.legendF))
allcolors <- c(color.legendF[sceF$CellType], color.legendA[sceA$CellType])
first.batch <- sce$Batch=="Nestorowa"
```

We now construct a plotting function for dimensionality reduction results, coloured by cell type:

```{r}
plotFUN <- function(fname, Y, subset=NULL, ..., xlab="tSNE 1",ylab="tSNE 2",main="") {
    if (is.null(subset)) {
        subset <- seq_len(nrow(Y))
    }
    png(fname,width=900,height=700)
    par(mfrow=c(1,1),mar=c(6,6,4,2),cex.axis=2,cex.main=3,cex.lab=2.5)
    plot(Y[,1], Y[,2], cex=2,
         pch=ifelse(first.batch, 21, 1)[subset], 
         col=ifelse(first.batch, "black", allcolors)[subset],
         bg=allcolors[subset], xlab=xlab, ylab=ylab, main=main) 
    dev.off()
}
```

... or by batch:


```{r}
batchcolor <- c("lavender","lightcoral")[first.batch + 1]
plotFUNb <- function(fname, Y, subset=NULL, ...) {
    if (is.null(subset)) {
        subset <- seq_len(nrow(Y))
    }
    png(fname,width=900,height=700)
    par(mfrow=c(1,1),mar=c(6,6,4,2),cex.axis=2,cex.main=3,cex.lab=2.5)
    plot(Y[,1], Y[,2], cex=2,
         pch=ifelse(first.batch, 21, 1)[subset], 
         col=ifelse(first.batch, "black", batchcolor)[subset],
         bg=batchcolor[subset], ...)
    dev.off()
}
```

Note that we'll only use the common cell types for PCA, just to provide a closer visualization of them.

```{r}
pca.retain <- sce$CellType %in% c("MEP", "GMP", "CMP") 
summary(pca.retain)
```

# No correction

Not doing any correction at all:

```{r}
t.unc <- as.matrix(t(logcounts(sce)))
```

Generating a t-SNE plot, using a high perplexity to capture the continuous nature of the population.
Note that this involves an internal PCA step with 50 dimensions, which matches the dimensionality reduction of `fastMNN()` and Seurat.

```{r}
set.seed(0)
library(BiocSingular)
pc.unc <- runPCA(t.unc, BSPARAM=IrlbaParam(), rank=50)
    
library(Rtsne)
tsne.unc <- Rtsne(pc.unc$x, perplexity = 90, pca=FALSE)
plotFUN("results/tsne_unc_type.png", tsne.unc$Y, main="Uncorrected", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_unc_batch.png", tsne.unc$Y, main="Uncorrected", xlab="tSNE 1",ylab="tSNE 2")
```

Generating a PCA plot.

```{r}
pca.unc <- runPCA(t.unc[pca.retain,], rank=2, BSPARAM=IrlbaParam())
plotFUN("results/pca_unc_type.png", pca.unc$x, subset=pca.retain, main="Uncorrected", xlab="PC 1",ylab="PC 2")
plotFUNb("results/pca_unc_batch.png", pca.unc$x, subset=pca.retain, main="Uncorrected", xlab="PC 1",ylab="PC 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.unc)
gc()
```

# Using classic MNN

We've turned down the sigma to improve mixing.

```{r}
mnn.out <- mnnCorrect(logcounts(sceF), logcounts(sceA), k=20, sigma=0.1)
t.mnn <- t(assay(mnn.out))
```

Generating a t-SNE plot.

```{r}
set.seed(0)
pc.mnn <- runPCA(t.mnn, BSPARAM=IrlbaParam(), rank=50)
tsne.mnn <- Rtsne(pc.mnn$x, perplexity = 90, pca=FALSE)
plotFUN("results/tsne_mnn_type.png", tsne.mnn$Y, main="MNN", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_mnn_batch.png", tsne.mnn$Y, main="MNN", xlab="tSNE 1",ylab="tSNE 2")
```

Generating a PCA plot.

```{r}
pca.mnn <- runPCA(t.mnn[pca.retain,], rank=2, BSPARAM=IrlbaParam())
plotFUN("results/pca_mnn_type.png", pca.mnn$x, subset=pca.retain, main="MNN", xlab="PC 1",ylab="PC 2")
plotFUNb("results/pca_mnn_batch.png", pca.mnn$x, subset=pca.retain, main="MNN", xlab="PC 1",ylab="PC 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.mnn)
gc()
```

# Using faster MNN

```{r}
set.seed(1000)
mnn.out2 <- fastMNN(logcounts(sceF), logcounts(sceA), k=20, 
    BSPARAM=IrlbaParam(deferred=TRUE))
t.mnn <- reducedDim(mnn.out2)
```

Generating a t-SNE plot - this already operates in low-dimensions, so there's no need for another PCA step.

```{r}
set.seed(0)
tsne.mnn <- Rtsne(t.mnn, perplexity = 90, pca=FALSE)
plotFUN("results/tsne_mnn2_type.png", tsne.mnn$Y, main="Fast MNN", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_mnn2_batch.png", tsne.mnn$Y, main="Fast MNN", xlab="tSNE 1",ylab="tSNE 2")
```

Generating a PCA plot.

```{r}
pca.mnn <- runPCA(t.mnn[pca.retain,], rank=2, BSPARAM=IrlbaParam())
plotFUN("results/pca_mnn2_type.png", pca.mnn$x, subset=pca.retain, main="Fast MNN", xlab="PC 1",ylab="PC 2")
plotFUNb("results/pca_mnn2_batch.png", pca.mnn$x, subset=pca.retain, main="Fast MNN", xlab="PC 1",ylab="PC 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.mnn)
gc()
```

# Using limma

Simply appling `removeBatchEffect` here:

```{r}
library(limma)
X.lm <- removeBatchEffect(as.matrix(logcounts(sce)), factor(first.batch))
t.lm <- t(X.lm)
```

Generating a t-SNE plot.

```{r}
set.seed(0)
pc.lm <- runPCA(t.lm, BSPARAM=IrlbaParam(), rank=50)
tsne.lm <- Rtsne(pc.lm$x, perplexity = 90, pca=FALSE)
plotFUN("results/tsne_limma_type.png", tsne.lm$Y, main="limma", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_limma_batch.png", tsne.lm$Y, main="limma", xlab="tSNE 1",ylab="tSNE 2")
```

Generating a PCA plot.

```{r}
pca.lm <- runPCA(t.lm[pca.retain,], rank=2, BSPARAM=IrlbaParam())
plotFUN("results/pca_limma_type.png", pca.lm$x, subset=pca.retain, main="limma", xlab="PC 1",ylab="PC 2")
plotFUNb("results/pca_limma_batch.png", pca.lm$x, subset=pca.retain, main="limma", xlab="PC 1",ylab="PC 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.lm)
gc()
```

# Using ComBat

```{r}
library(sva)
Z <- factor(first.batch)
X.combat <- ComBat(as.matrix(logcounts(sce)), Z, mod=NULL,prior.plots = FALSE)
t.combat <- t(X.combat)
```

Generating a t-SNE plot.

```{r}
set.seed(0)
pc.combat <- runPCA(t.combat, BSPARAM=IrlbaParam(), rank=50)
tsne.combat <- Rtsne(pc.combat$x, perplexity = 90, pca=FALSE)
plotFUN("results/tsne_combat_type.png", tsne.combat$Y, main="ComBat", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_combat_batch.png", tsne.combat$Y, main="ComBat", xlab="tSNE 1",ylab="tSNE 2")
```

Generating a PCA plot.

```{r}
pca.combat <- runPCA(t.combat[pca.retain,], rank=2, BSPARAM=IrlbaParam())
plotFUN("results/pca_combat_type.png", pca.combat$x, subset=pca.retain, main="ComBat", xlab="PC 1",ylab="PC 2")
plotFUNb("results/pca_combat_batch.png", pca.combat$x, subset=pca.retain, main="ComBat", xlab="PC 1",ylab="PC 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.combat)
gc()
```

# Using CCA from `r CRANpkg("Seurat")`

## Using our normalized values

This requires some effort to set up all the objects without funnelling it through `r CRANpkg("Seurat")`'s pipeline.
Bypassing their pipeline is deliberate to ensure that we are comparing only the effect of correction, not the preprocessing.

```{r}
library(Seurat)
SeuA <- CreateSeuratObject(logcounts(sceA))
SeuF <- CreateSeuratObject(logcounts(sceF))
SeuA@meta.data$group <- "group1"
SeuF@meta.data$group <- "group2"

SeuA <- ScaleData(SeuA)
SeuF <- ScaleData(SeuF)
Y <- RunCCA(SeuF, SeuA, genes.use=rownames(sceA), do.normalize=FALSE)
suppressWarnings(Y <- AlignSubspace(Y, grouping.var="group", dims.align=1:20))

t.cca <- Y@dr$cca.aligned@cell.embeddings
```

Generating a t-SNE plot - again, we're working in low-dimensional space already, so no need for a PCA beforehand.

```{r}
set.seed(0)
tsne.cca <- Rtsne(t.cca, perplexity = 90, pca=FALSE)
plotFUN("results/tsne_cca_type.png", tsne.cca$Y, main="CCA", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_cca_batch.png", tsne.cca$Y, main="CCA", xlab="tSNE 1",ylab="tSNE 2")
```

Generating a PCA plot.

```{r}
pca.cca <- runPCA(t.cca[pca.retain,], rank=2, BSPARAM=IrlbaParam())
plotFUN("results/pca_cca_type.png", pca.cca$x, subset=pca.retain, main="CCA", xlab="PC 1",ylab="PC 2")
plotFUNb("results/pca_cca_batch.png", pca.cca$x, subset=pca.retain, main="CCA", xlab="PC 1",ylab="PC 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.cca)
gc()
```

## Using their normalized values

Or we just give in and use their values.

```{r}
SeuA <- CreateSeuratObject(counts(sceA))
SeuF <- CreateSeuratObject(counts(sceF))
SeuA@meta.data$group <- "group1"
SeuF@meta.data$group <- "group2"

SeuA <- NormalizeData(SeuA)
SeuF <- NormalizeData(SeuF)
SeuA <- ScaleData(SeuA)
SeuF <- ScaleData(SeuF)
Y <- RunCCA(SeuF, SeuA, genes.use=rownames(sceA))
suppressWarnings(Y <- AlignSubspace(Y, grouping.var="group", dims.align=1:20))

t.cca <- Y@dr$cca.aligned@cell.embeddings
```

Generating a t-SNE plot.

```{r}
set.seed(0)
tsne.cca2 <- Rtsne(t.cca, perplexity = 90, pca=FALSE)
plotFUN("results/tsne_cca2_type.png", tsne.cca2$Y, main="CCA native", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_cca2_batch.png", tsne.cca2$Y, main="CCA native", xlab="tSNE 1",ylab="tSNE 2")
```

Generating a PCA plot.

```{r}
pca.cca2 <- runPCA(t.cca[pca.retain,], rank=2, BSPARAM=IrlbaParam())
plotFUN("results/pca_cca2_type.png", pca.cca2$x, subset=pca.retain, main="CCA native", xlab="PC 1",ylab="PC 2")
plotFUNb("results/pca_cca2_batch.png", pca.cca2$x, subset=pca.retain, main="CCA native", xlab="PC 1",ylab="PC 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.cca)
gc()
```

# Finishing up

Making the legend (using PDF for better resolution).

```{r}
pdf(file="results/legend.pdf", width=7, height=7)
plot(0,0,type="n", bty="n", axes=FALSE, xlab="", ylab="")
legend(x=-1, y=1, legend=names(color.legendF), pch=21, cex=2.5, col="black", pt.bg=color.legendF, bty="n")
legend(x=0, y=1, legend=names(color.legendA)[1:3], pch=21, cex=2.5, col="black", pt.bg=color.legendA[1:3], bty="n")
dev.off()
```

Printing the session information.

```{r}
sessionInfo()
```

