---
title: "Data preparation for the haematopoiesis analysis"
author: "Aaron Lun, based on code by Laleh Haghverdi and Michael Morgan"
date: "Revised: 27 February 2019"
output:
  BiocStyle::html_document:
    toc_float: true
    fig_caption: no
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview

This script prepares data for the haematopoiesis batch correction analysis.
We'll be storing data using the `r Biocpkg("BiocFileCache")` package to avoid re-downloading files more than once.
We also set the seed in preparation for any randomized algorithms.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache(ask=FALSE)    
set.seed(123456)
```

# Processing the Nestorowa data set

We read in the counts from the [Nestorowa _et al._ (2016)](https://doi.org/10.1182/blood-2016-05-716480) study.
We use `readSparseCounts` to preserve the sparsity of the data:

```{r}
library(scater)
fname.F <- bfcrpath(bfc, 
    "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81682/suppl/GSE81682_HTSeq_counts.txt.gz")
dataF <- readSparseCounts(fname.F)
```

We construct a `SingleCellExperiment` object:

```{r}
library(SingleCellExperiment)
sceF <- SingleCellExperiment(list(counts=dataF))
sceF <- sceF[!grepl("^_", rownames(sceF)),]
sceF <- sceF[!grepl("^ERCC-", rownames(sceF)),] # removing spike-ins completely.
dim(sceF)
```

We load in the metadata to specify the type of each cell.

```{r}
meta.fname <- bfcrpath(bfc, 
    "http://blood.stemcells.cam.ac.uk/data/all_cell_types.txt")
metaF <- read.table(meta.fname, stringsAsFactors = FALSE, header=TRUE, check.names=FALSE)
metainds <- match(colnames(dataF), rownames(metaF))

# This will contain NA's... which is okay, at this point, to preserve length.
metaF <- metaF[metainds,] 
head(metaF)
```

We actually define the cell type, which requires some work as some cells have multiple annotations.

```{r}
metatypeF <- rep("other", nrow(metaF))
for (col in rev(colnames(metaF))) { # reverse, so earlier columns end up overwriting later ones.
    chosen <- metaF[,col]==1
    metatypeF[chosen] <- sub("[0-9]?_.*", "", col)
}
metatypeF[metatypeF=="ESLAM"] <- "HSPC"
```

If cell type metadata is missing, we fill it in using information from the cell sorting label.

```{r}
missing.meta <- is.na(metainds)
metatypeF[missing.meta] <- sub("_.*", "", colnames(dataF)[missing.meta])
metatypeF[metatypeF=="LT-HSC"] <- "LTHSC"
metatypeF[metatypeF=="Prog"] <- "other"
```

This is added to our `SingleCellExperiment` object:

```{r}
sceF$CellType <- metatypeF
table(metatypeF)
```

We perform size factor normalization within this data set.
Preclustering:

```{r}
library(scran)
library(BiocSingular)
clustF <- quickCluster(sceF, use.ranks=FALSE, BSPARAM=IrlbaParam())
table(clustF) 
```

... and computing the size factors per cell:

```{r}
sceF <- computeSumFactors(sceF, cluster=clustF)
summary(sizeFactors(sceF))    
plot(Matrix::colSums(counts(sceF)), sizeFactors(sceF), log="xy",
    xlab="Library sizes", ylab="Size factors")
```

```{r, echo=FALSE, results="hide"}
# Cleaning up memory.
gc() 
```

# Processing the Paul data set

We read in the counts from the [Paul _et al._ (2015)](https://doi.org/10.1016/j.cell.2015.11.013) study.

```{r}
fname.A <- bfcrpath(bfc, 
    "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE72nnn/GSE72857/suppl/GSE72857_umitab.txt.gz")
dataA <- readSparseCounts(fname.A, quote='"')
sceA <- SingleCellExperiment(list(counts=dataA))
dim(sceA)
```

We managed to get some metadata about cell types directly from the authors.
We subset to those cells that are in the metadata:

```{r}
map.fname <- bfcrpath(bfc, 
    "https://raw.githubusercontent.com/MarioniLab/MNN2017/master/Haematopoiesis/MAP.csv")
metaA <- read.csv2(map.fname, sep=",", header=FALSE, row.names=1)
metainds <- match(rownames(metaA), colnames(sceA))
sceA <- sceA[,metainds]
ncol(sceA)
```

We organize the cell type labels:

```{r}
metatypeA <- character(nrow(metaA))
metatypeA[metaA[,1]<7] <- "ERY"
metatypeA[metaA[,1]>6 & metaA[,1]<12] <- "CMP"
metatypeA[metaA[,1]>11] <- "GMP"
sceA$CellType <- metatypeA
table(metatypeA)
```

And we perform size factor normalization within this data set.

```{r}
clustA <- quickCluster(sceA, use.ranks=FALSE, BSPARAM=IrlbaParam())
table(clustA) 
sceA <- computeSumFactors(sceA, cluster=clustA, min.mean=0.1)
summary(sizeFactors(sceA))    
plot(Matrix::colSums(counts(sceA)), sizeFactors(sceA), log="xy",
    xlab="Library sizes", ylab="Size factors")
```

```{r, echo=FALSE, results="hide"}
# Cleaning up memory.
gc()
```

# Standardizing gene sets across batches

The Paul data set uses a rather inconvenient annotation, so this requires some work to determine their Ensembl IDs.

```{r}
library(org.Mm.eg.db)
anno <- select(org.Mm.eg.db, keys=keys(org.Mm.eg.db),
    keytype="ENTREZID", column=c("ENSEMBL", "SYMBOL"))

# Determining which subnames exist in the annotation.
individual.names <- strsplit(rownames(sceA), ";")
row.number <- rep(seq_along(individual.names), lengths(individual.names))
individual.names <- unlist(individual.names)    

ens.id <- anno$ENSEMBL[match(individual.names, anno$SYMBOL)]
has.ens <- !is.na(ens.id)
ens.id <- ens.id[has.ens]
row.number <- row.number[has.ens]

# Assigning them back to the original rows.
final.name <- ens.id[match(seq_len(nrow(sceA)), row.number)]
head(final.name)
```

We discard rows without annotation.

```{r}
keep <- !is.na(final.name)
sceA <- sceA[keep,]
rownames(sceA) <- final.name[keep]
summary(keep)
```

# Finishing up

We save the datasets to file:

```{r}
saveRDS(sceA, file="haem_data_A.rds")
saveRDS(sceF, file="haem_data_F.rds")
```

... and we report the session information for posterity:

```{r}
sessionInfo()
``` 
