---
title: "Batch correction for the pancreas data"
author: "Aaron Lun, based on code by Laleh Haghverdi and Michael Morgan"
date: "Revised: 27 February 2019"
output:
  BiocStyle::html_document:
    toc_float: true
    fig_caption: no
---


```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Overview    

This script performs batch correction with each method and creates t-SNE plots for the pancreas data sets.
First, we read in the various `SingleCellExperiment` objects:

```{r}
library(SingleCellExperiment)
dir.create("results", showWarning=FALSE)
sceA <- readRDS("sce.gse81076.rds")
sceB <- readRDS("sce.gse85241.rds")
sceC <- readRDS("sce.gse86473.rds")
sceD <- readRDS("sce.emtab5601.rds")
```

We also read in the HVG results.

```{r}
decA <- readRDS("dec.gse81076.rds")
decB <- readRDS("dec.gse85241.rds")
decC <- readRDS("dec.gse86473.rds")
decD <- readRDS("dec.emtab5601.rds")
```

We define the universe of genes that are common across all batches.

```{r}
universe <- Reduce(intersect, 
	list(rownames(decA), rownames(decB), rownames(decC), rownames(decD)))
universe <- universe[!grepl("^ERCC-", universe)] # removing spike-ins.
length(universe)
```

We adjust the scale of the batches to minimize differences in variance.

```{r}
library(batchelor)
nout <- multiBatchNorm(sceA[universe,], sceB[universe,], 
    sceC[universe,], sceD[universe,])
sceA <- nout[[1]]
sceB <- nout[[2]]
sceC <- nout[[3]]
sceD <- nout[[4]]
```

We keep all genes with average biological components (across all batches) greater than zero.

```{r}
combined.bio <- scran::combineVar(decA[universe,], decB[universe,], 
    decC[universe,], decD[universe,], weighted=FALSE)
chosen <- universe[combined.bio$bio > 0]
length(chosen)
sceA <- sceA[chosen,]
sceB <- sceB[chosen,]
sceC <- sceC[chosen,]
sceD <- sceD[chosen,]
```

# Setting up the visualizations

We organize the cell type colors.

```{r}
cell.types <- c(rep("unknown", ncol(sceA)), rep("unknown", ncol(sceB)), sceC$CellType, sceD$CellType)
type.colors <- c(
    Acinar="#ffff00",
    Alpha="#ff0000",
    Beta="#c400ff",
    Delta="#ff7800",
    Ductal="#00f5ff",
    Gamma="#0000ff",
    Other="#000000",
    unknown="grey80"
)
cell.types[!cell.types %in% names(type.colors)] <- "Other"
allcolors <- type.colors[cell.types]
```

We organize the batch colors.

```{r}
batch.id <- rep(1:4, c(ncol(sceA), ncol(sceB), ncol(sceC), ncol(sceD)))
batchcolor <- c("lavender", "lightcoral", "goldenrod1", "lightblue")[batch.id]
```

We create a plotting function based on cell type.

```{r}
plotFUN <- function(fname, Y, subset=NULL, ..., xlab="tSNE 1",ylab="tSNE 2",main="") {
    if (is.null(subset)) {
        subset <- seq_len(nrow(Y))
    }
    png(fname,width=900,height=700)
    par(mfrow=c(1,1),mar=c(6,6,4,2),cex.axis=2,cex.main=3,cex.lab=2.5)
    plot(Y[,1], Y[,2], cex=2, pch=21, bg=allcolors[subset], xlab=xlab, ylab=ylab, main=main) 
    dev.off()
}
```

... and another one that colors by batch.

```{r}
plotFUNb <- function(fname, Y, subset=NULL, ...) {
    if (is.null(subset)) {
        subset <- seq_len(nrow(Y))
    }
    png(fname,width=900,height=700)
    par(mfrow=c(1,1),mar=c(6,6,4,2),cex.axis=2,cex.main=3,cex.lab=2.5)
    plot(Y[,1], Y[,2], cex=2, pch=21, bg=batchcolor[subset], ...)
    dev.off()
}
```

# No correction

Doing no correction.

```{r}
combined <- cbind(logcounts(sceA), logcounts(sceB), logcounts(sceC), logcounts(sceD))
t.unc <- as.matrix(t(combined))
```

Generating a t-SNE plot.

```{r}
set.seed(0)
library(BiocSingular)
pc.unc <- runPCA(t.unc, rank=50, BSPARAM=IrlbaParam())
    
library(Rtsne)
tsne.unc <- Rtsne(pc.unc$x, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_unc_type.png", tsne.unc$Y, main="Uncorrected", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_unc_batch.png", tsne.unc$Y, main="Uncorrected", xlab="tSNE 1",ylab="tSNE 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.unc)
gc()
```

# Using classic MNN

We've turned down the sigma to improve mixing.

```{r}
mnn.out <- mnnCorrect(logcounts(sceA), logcounts(sceB), 
    logcounts(sceC), logcounts(sceD), k=20, sigma=0.1)
t.mnn <- as.matrix(t(do.call(cbind, mnn.out$corrected)))
```

Generating a t-SNE plot.

```{r}
set.seed(0)
pc.mnn <- runPCA(t.mnn, rank=50, BSPARAM=IrlbaParam())
tsne.mnn <- Rtsne(pc.mnn$x, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_mnn_type.png", tsne.mnn$Y, main="MNN", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_mnn_batch.png", tsne.mnn$Y, main="MNN", xlab="tSNE 1",ylab="tSNE 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.mnn)
gc()
```

# Using faster MNN

Here we use a hierarchical merge, within the protocols first and then across protocols.

```{r}
set.seed(1000)
pcs <- multiBatchPCA(
    cosineNorm(logcounts(sceA)), 
    cosineNorm(logcounts(sceB)), 
    cosineNorm(logcounts(sceC)), 
    cosineNorm(logcounts(sceD)),
    BSPARAM=IrlbaParam(deferred=TRUE)
)
mnn.out.umi <- fastMNN(pcs[[1]], pcs[[2]], k=20, pc.input=TRUE)
mnn.out.ss2 <- fastMNN(pcs[[3]], pcs[[4]], k=20, pc.input=TRUE)
mnn.out2 <- fastMNN(mnn.out.umi$corrected, mnn.out.ss2$corrected, k=20, pc.input=TRUE)
t.mnn <- mnn.out2$corrected
```

Generating a t-SNE plot (no need for another PCA step, as we're already working in low dimensions.

```{r}
set.seed(0)
tsne.mnn <- Rtsne(t.mnn, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_mnn2_type.png", tsne.mnn$Y, main="Fast MNN", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_mnn2_batch.png", tsne.mnn$Y, main="Fast MNN", xlab="tSNE 1",ylab="tSNE 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.mnn)
gc()
```

# Using `r Biocpkg("limma")`

Using `removeBatchEffect` directly.

```{r}
library(limma)
X.lm <- removeBatchEffect(as.matrix(combined), factor(batch.id))
t.lm <- t(X.lm)
```

Generating a t-SNE plot.

```{r}
set.seed(0)
pc.lm <- runPCA(t.lm, rank=50, BSPARAM=IrlbaParam())
tsne.lm <- Rtsne(pc.lm$x, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_limma_type.png", tsne.lm$Y, main="limma", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_limma_batch.png", tsne.lm$Y, main="limma", xlab="tSNE 1",ylab="tSNE 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.lm)
gc()
```

# Using `ComBat`

As before:

```{r}
library(sva)
Z <- factor(batch.id)
X.combat <- ComBat(as.matrix(combined), Z, mod=NULL,prior.plots = FALSE)
t.combat <- t(X.combat)
```

Generating a t-SNE plot.

```{r}
set.seed(0)
pc.com <- runPCA(t.combat, rank=50, BSPARAM=IrlbaParam())    
tsne.combat <- Rtsne(pc.com$x, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_combat_type.png", tsne.combat$Y, main="ComBat", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_combat_batch.png", tsne.combat$Y, main="ComBat", xlab="tSNE 1",ylab="tSNE 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.combat)
gc()
```

# Using CCA

## Based on the pre-processed data

Only using the two batches with the cell type labels, as I don't know how to do multiple corrections.

```{r}
library(Seurat)
SeuC <- CreateSeuratObject(logcounts(sceC))
SeuD <- CreateSeuratObject(logcounts(sceD))
SeuC@meta.data$group <- "group1"
SeuD@meta.data$group <- "group2"

SeuC <- ScaleData(SeuC)
SeuD <- ScaleData(SeuD)
Y <- RunCCA(SeuC, SeuD, genes.use=rownames(sceA), do.normalize=FALSE)
suppressWarnings(Y <- AlignSubspace(Y, grouping.var="group", dims.align=1:20))

t.cca <- Y@dr$cca.aligned@cell.embeddings
```

Generating a t-SNE plot - again, we're already in low-dimensional space, so no need to do more.

```{r}
set.seed(0)
tsne.cca <- Rtsne(t.cca, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_cca_type.png", tsne.cca$Y, subset=(batch.id%in%c(3,4)), main="CCA", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_cca_batch.png", tsne.cca$Y, subset=(batch.id %in% c(3,4)), main="CCA", xlab="tSNE 1",ylab="tSNE 2")
```

## Normalizing it their way

Repeating this using their normalization pipeline.

```{r}
SeuC <- CreateSeuratObject(counts(sceC))
SeuD <- CreateSeuratObject(counts(sceD))
SeuC@meta.data$group <- "group1"
SeuD@meta.data$group <- "group2"

SeuC <- NormalizeData(SeuC)
SeuD <- NormalizeData(SeuD)
SeuC <- ScaleData(SeuC)
SeuD <- ScaleData(SeuD)
Y <- RunCCA(SeuC, SeuD, genes.use=rownames(sceA))
suppressWarnings(Y <- AlignSubspace(Y, grouping.var="group", dims.align=1:20))

t.cca <- Y@dr$cca.aligned@cell.embeddings
```

Generating a t-SNE plot.

```{r}
set.seed(0)
tsne.cca2 <- Rtsne(t.cca, perplexity = 30, pca=FALSE)
plotFUN("results/tsne_cca2_type.png", tsne.cca2$Y, subset=(batch.id%in%c(3,4)), main="CCA native", xlab="tSNE 1",ylab="tSNE 2")
plotFUNb("results/tsne_cca2_batch.png", tsne.cca2$Y, subset=(batch.id%in%c(3,4)), main="CCA native", xlab="tSNE 1",ylab="tSNE 2")
```

```{r, echo=FALSE, results="hide"}
rm(t.cca)
gc()
```

# Session information

```{r}
sessionInfo()
```    
